---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Imports

```{r}
library(zoo)
library(data.table)
library(dplyr)
library(lubridate)
library(ggplot2)
library(knitr)
```

## Loading and preprocessing the data

```{r}
if(!file.exists("activity.csv")) {
    if(file.exists("activity.zip")) {
        unzip("activity.zip")
    } else {
        stop("Cannot find data file")
    }
}
decodeTimeIdentifier <- function(timeIdentifier) {
    mnts <- timeIdentifier %% 100
    hrs <- (timeIdentifier - mnts) / 100
    hrs + mnts / 60
}
stepsData <- fread("activity.csv")    
stepsData <- mutate(stepsData,date = ymd(date), hour = decodeTimeIdentifier(interval),timestamp = date + hour * 3600)
```


## What is mean total number of steps taken per day?

A histogram of the total number of steps per day is as follows:

```{r}
dailySteps <- stepsData %>% group_by(date) %>% summarize(steps = sum(steps,na.rm=TRUE))
hist(dailySteps$steps,main="Daily Steps",xlab="Total steps every day")
```

The following is the mean total number of steps taken per day:

```{r}
meanDailySteps <- mean(dailySteps$steps)
round(meanDailySteps)
```

The following is the median total number of steps taken per day:

```{r}
medianDailySteps<-median(dailySteps$steps)
```


## What is the average daily activity pattern?

```{r}
activityPattern <- stepsData %>% group_by(hour) %>% summarize(steps = mean(steps,na.rm=TRUE))
plot(activityPattern$hour,activityPattern$steps,type="l",xlab="The hour of the day",ylab="Steps",main="Daily Activity Pattern")
```

The 5-minute intervals with the most average number of steps is:

```{r}
activityPattern %>% filter(steps == max(steps))
```

## Imputing missing values

The number of records with missing values are:

```{r}
stepsData %>% filter(is.na(steps)) %>% nrow()
```

We'll use the *carry-forward* method of filling missing values. The 5-minute interval with an `NA` steps will use the `steps` value of the interval that came before it. Furthermore we'll remove the first day reading of the data set since that entire day consists of all `NA` values and we assume that data collection hasn't been started from that day.

```{r}
stepsDataImputed <- filter(stepsData,date != ymd("2012-10-01")) %>% mutate(steps=na.locf(steps))
```


A histogram of total steps every day

```{r}
dailyStepsImputed <- stepsDataImputed %>% group_by(date) %>% summarize(steps = sum(steps,na.rm=TRUE))
hist(dailyStepsImputed$steps,main="Daily Steps",xlab="Total steps every day (imputed)")
```

The following is the mean total number of steps taken per day from the imputed data:

```{r}
meanDailyStepsImputed <- mean(dailyStepsImputed$steps)
round(meanDailyStepsImputed)
```

The following is the median total number of steps taken per day from the imputed data:

```{r}
medianDailyStepsImputed <- median(dailyStepsImputed$steps)
```

Apparently imputing values introduces a small difference of the data characteristics, but the difference is not significant (about one percent).

```{r}
imputingSummary <- data.frame(
    Original=c(round(meanDailySteps),medianDailySteps),
    Imputed=c(round(meanDailyStepsImputed),medianDailyStepsImputed),
    Difference=c(
        sprintf("%0.2f%%",abs(meanDailyStepsImputed-meanDailySteps)/meanDailySteps*100),
        sprintf("%0.2f%%",abs(medianDailyStepsImputed-medianDailySteps)/medianDailySteps*100)
        ),
    row.names=c("Mean","Median")
    )
kable(imputingSummary)
```


## Are there differences in activity patterns between weekdays and weekends?

```{r}
activityPatternByDayType <- stepsData %>% 
    mutate(dayType=ifelse(weekdays(date) %in% c("Saturday","Sunday"),"Weekend","Weekday")) %>% 
    group_by(dayType,hour) %>% summarize(steps = mean(steps,na.rm=TRUE))
ggplot(activityPatternByDayType) + 
    aes(hour,steps) + 
    geom_line() +
    coord_cartesian(xlim=c(0,24)) +
    ylab("Average Number of Steps") + xlab("Hour of the Day") + 
    ggtitle("Weekday vs Weekend Activity Pattern") +
    facet_grid(dayType ~ .)
    
```

